---
title: "MXN600 Major Data Anaysis Project"
author: "Joanna Salerno, Pavan Asopa, Sevin Nejadi, Fernanda Martins Giuriati"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The credit risk models our lending start up company uses are of the utmost importance to the functioning and ultimate success of the company. As we were recently acquired by a regional Australian bank, the success of our company also impacts the success of the bank. Recently, some of the bank's senior financial analysts have raised concerns about the credit risk models we have been using. They reviewed our models' performance benchmarks and feel as though our models are not suitable for use in a setting in which they are subject to strict regulatory requirements.

Thus, we have been tasked by the bank's management to rebuild our credit risk model from the ground up. As per management, our main objective is to use information known at the time of a loan application to build a model that predicts loan default. We will follow a standard statistical analysis process, which will be guided by the following questions:

1. How does this new model perform compared to the one used previously? How can it be expected to perform on new loan applications?
2. What are the important variables in this model and how do they compare to variables that are traditionally important for predicting credit risk in the banking sector?

Furthermore, management has consulted with an expert statistician, who has suggested we also account for variation in trends that may exist either between different jurisdictions or over time. The following questions will guide this second part of our analysis:

3. Can accounting for this variation (e.g., state/zip-code and time) improve performance benchmarks?
4. Are there any surprising differences in variables that are important for predicting credit risk?
5. Does credit risk change over time or between states? This is not something the bank has previously
investigated and results may inform modified loan policies in the future.

This report will document our entire analysis process, beginning with data exploration and cleaning, to model building and interpretations of our results.

# Setup

We will first load in the required libraries for our data exploration and analysis process.

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(MASS)
library(GGally)
library(lme4)
library(lmerTest)
library(DHARMa)
library(pROC)
library(dplyr)
```

Next, we will load in our datasets. We have a total of 4:

(1) A training dataset that we will use to build and train our model
(2) A test dataset that we will use to test the fit of our model(s)
(3) A validation dataset that we will use to assess the performance of our model(s)
(4) An extended dataset that includes the necessary variables for us to account for variation such as location and time

```{r load_data}
train_data <- read.csv("benchmark_training_loan_data.csv")
test_data <- read.csv("benchmark_testing_loan_data.csv")
val_data <- read.csv("benchmark_validation_loan_data.csv")
extended <- read.csv("extendend_version_loan_data.csv")
```

# Exploratory Analysis

We will begin by exploring the available data to understand how each variable is distributed and to identify any potential data quality issues. We will also investigate the relationships between the different variables to see whether any variables are highly correlated with one another.

Note: For this exploration portion of our analysis, we will be using the training dataset.

We will first explore the training dataset to understand its structure and the variables it is comprised of.

```{r head_data}
head(train_data)
```

At first glance of the first 6 rows of this dataset, we notice there is a value of n/a in the employment length column. There are also a few zero values in a few columns. This will prompt us to further explore the data for any true missing values.

```{r data_dim}
dim(train_data)
```


The training dataset contains a total of 23,052 observations of 19 variables.

```{r data_structure}
str(train_data)
```

Upon further investigation of the structure of the training data, we can see that 14 of the 19 variables are currently of numeric type, and the remaining 5 variables are characters.

We will now remove variable X, which is the number of each record, from all datasets, as it is not necessayr for this analysis.

```{r remove_X_from_datasets}
train_data <- train_data[, -1]
val_data <- val_data[, -1]
test_data <- test_data[, -1]
extended <- extended[, -1]
```

We will now further investigate each variable to determine whether there is any missing data or outliers.

```{r data_summary}
summary(train_data)
```

Included above is the numeric distributions of each of the included variables. Based on the above summary, it appears as though there may exist one or multiple outliers in a few variables: annual income, inquiries in the last 6 months, open accounts, revolving balances, and total accounts. However, we will need to further investigate the distribution of all variables to better visualize this to determine whether these actually appear to be outliers.

Before producing some exploratory plots, we will briefly explore the data to see whether there are missing values for us to handle.

```{r sum_na}
colSums(is.na(train_data))
```

Based on the above output, it appears as though this dataset does not contain any missing data in the form of NA values. Now, we will explore the n/a values present in the columns which include string data.

```{r string_na_sum}
sum(train_data == 'n/a')
```

```{r string_na_rows}
na_rows <- train_data %>% filter_all(any_vars(. %in% c('n/a')))
head(na_rows)
```

Based on the above output, there are a total of 591 n/a (string) in this dataset. These all appear to be from the employment length column. We will move forward and assume that this is not truly missing data, but rather, means that the applicant is not currently employed. We will further examine the ditribution of each variable by creating exploratory plots, and this will further clarify the meaning of some of the values contained within each column of the dataset.

## Exploratory Plots

Before we can create exploratory plots, there are a few variables we will need to convert to factors. This will help us in creating our visualizations as well as conducting our analyses, so we can consider the selected variables as factors with different levels rather than simply strings.

```{r convert_vars_to_factors_train}
train_data$term <- as.factor(train_data$term)
train_data$emp_length <- as.factor(train_data$emp_length)
train_data$home_ownership <- as.factor(train_data$home_ownership)
train_data$verification_status <- as.factor(train_data$verification_status)
train_data$purpose <- as.factor(train_data$purpose)
train_data$repay_fail <- as.factor(train_data$repay_fail)
```

```{r}
test_data$term <- as.factor(test_data$term)
test_data$emp_length <- as.factor(test_data$emp_length)
test_data$home_ownership <- as.factor(test_data$home_ownership)
test_data$verification_status <- as.factor(test_data$verification_status)
test_data$purpose <- as.factor(test_data$purpose)
test_data$repay_fail <- as.factor(test_data$repay_fail)
```


```{r convert_vars_to_factors_val}
val_data$term <- as.factor(val_data$term)
val_data$emp_length <- as.factor(val_data$emp_length)
val_data$home_ownership <- as.factor(val_data$home_ownership)
val_data$verification_status <- as.factor(val_data$verification_status)
val_data$purpose <- as.factor(val_data$purpose)
val_data$repay_fail <- as.factor(val_data$repay_fail)
```

```{r check_imbalance_data}
plot(train_data$repay_fail)
```

Despite the significant class imbalance in the dataset, logistic regression still maintains its probabilistic nature. However, it's important to note that imbalanced data can affect the model's predictive performance, potentially leading to a bias towards the majority class.

Now, we will create some histograms to visualize the distributions of some of the string/character variables in this dataset.

```{r histograms}
p1 <- ggplot(data = train_data, aes(int_rate, fill = repay_fail)) +
  geom_histogram(binwidth=5)

p2 <- ggplot(data = train_data, mapping = aes(x = verification_status, y = annual_inc)) +
  geom_boxplot() +
  theme_bw()

p3 <- ggplot(data = train_data, mapping = aes(x = repay_fail, y = loan_amnt)) +
  geom_boxplot() +
  theme_bw()

p4 <- ggplot(data = train_data, mapping = aes(x = repay_fail, y = annual_inc)) +
  geom_boxplot() +
  theme_bw()

p5 <- ggplot(data = train_data, mapping = aes(x = annual_inc, y = loan_amnt)) +
  geom_point() +
  theme_bw()

ggarrange(p1, p2)
ggarrange(p3, p4)
ggarrange(p5)
```

Explanations: 

Plot1 is a histogram that displays the distribution of interest rates in the dataset. The bars in the histogram are filled based on whether the loan was repaid successfully or not, where fill colours represent the two categories 0 for no failure and 1 for failure of loan repayment. Based on the plot, a trend can be seen that as the interest rate increases so does the chances of loan repayment failures.

Plot2 is a boxplot that shows the distribution of annual incomes based on different verification statuses. Each box represent a category of verification status. Based on the plot, we can see that we have similar sized boxes for all three categories which suggests that the median income for applicants in these categories is relatively consistent. However, we can see the presence of outlier in all three verification_status categories indicating that there are individuals with very high income within each group.

Plot3 is a boxplot that compares the loan amounts for both successfull and unsuccessfull loan repayments. Based on the plot, it can be seen that the median loan amount for both categories of loan repayment status are more or less consistent. The presence of outlier in both categories indicates that there are individuals with very high loan amount within each group. 

Plot4 is a boxplot that compares annual income between successful and unsuccessful loan repayments. Based on the plot it can be seen that most individuals that successfully paid off their loans are more or less consistent with the ones who failed with an exception of ouliers in both categories.

Plot5 is scatter plot that displays relationship annual income and loan amount. 

Here we create some distribution plots to have a better visualization of the data:

```{r dist_plots_1}
numeric_vars <- c("loan_amnt", "annual_inc", "dti", "delinq_2yrs", "inq_last_6mths",
                  "open_acc", "pub_rec", "revol_util", "revol_bal", "total_acc",
                  "credit_age_yrs")
par(mfrow=c(3,4))
for (var in numeric_vars) {
  hist(train_data[[var]], main=var, xlab=var)
}
```

The above plot displays the distribution of all numeric variables in the dataset in the form of a histogram. 

```{r dist_plots_2}
ggplot(train_data, aes(x = purpose)) +
  geom_bar() +
  labs(x = "Purpose", y = "Frequency") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The above bar plot represents the frequency of each unique value in the purpose variable. From the plot it can be seen that the most common reasons to take loans are debt_consolidation, credit_card, home_improvement and car.

```{r dist_plots_3}
ggplot(train_data, aes(x = term, y = loan_amnt)) +
  geom_boxplot() +
  labs(x = "Term", y = "Loan Amount")
```

The above plot consists of two side-by-side boxplots, one for each loan term category. Each boxplot provides information about the distribution of loan amounts for its respective term category. Based on the plot, we can see that, on average, loans with 60 months term have higher loan amounts than those with a 36 month term which is an expected behaviour. Also, we can observe some outliers in the 36 months term which indicates that there are a few loan amounts significantly higher than the typical range of loan amounts for that category.

Now, we will create a plot that depicts the correlation between the different numeric variables; it includes both correlation coefficients and visuals of the distribution of each included variable.

```{r ggpairs_plot}
# create ggpairs plot with the specified columns
ggpairs_plot <- ggpairs(train_data[, c("loan_amnt","int_rate", "annual_inc", "dti",
                                      "delinq_2yrs", "inq_last_6mths", "open_acc",
                                      "pub_rec", "revol_util", "revol_bal", "total_acc",
                                      "credit_age_yrs")],
                        progress = FALSE)

#ggpairs_plot
```

Below, we create a correlation matrix of the numeric variables to help us understand the relationship between the different numeric variables.

```{r corr_matrix}
columns <- c("loan_amnt", "annual_inc", "dti", "delinq_2yrs", "inq_last_6mths", "open_acc", "pub_rec", "revol_util", "revol_bal", "total_acc", "credit_age_yrs","int_rate")
corr_matrix <- cor(train_data[columns], method = "pearson")
ggcorr(corr_matrix, hjust = 1, size = 3, color = "grey")
```

The correlation matrix (heatmap) reveals several noteworthy patterns:

Upon reviewing the correlation matrix, it becomes evident that certain features exhibit a high degree of positive correlation, such as "total_acc" and "open_acc." Additionally, "credit_age_yrs" and "total_acc," as well as "revol_util" and "int_rate," while not displaying correlations as strong as those between "total_acc" and "open_acc," still exhibit a notable level of correlation.

Furthermore, following our examination of this matrix, each variable within pairs of highly correlated predictors will be analyzed while fitting models with the goal of identifying and refining the best model.

# New Credit Risk Model

Keeping our above exploratory analysis in mind, as well as reviewing the [Five Cs of Credit](https://www.investopedia.com/ask/answers/040115/what-most-important-c-five-cs-credit.asp) information, we have decided that the below variables are important predictors in assessing the risk of credit default:

annual_inc, emp_length, verification_status, dti, credit_age_yrs, home_ownership, total_acc, revol_bal, revol_util, int_rate, term, purpose, loan_amnt, inq_last_6mths, delinq_2yrs, pub_rec
                   
Now, we'll move on to building our first model. We'll first need to standardize our numeric variables so that they can be accurately compared:

```{r train_standardisation}
numeric_columns <- sapply(train_data, is.numeric)  # Identify numeric columns

# Scale the numeric columns while keeping column names
train_data[, numeric_columns] <- scale(train_data[, numeric_columns])

train_scaled_data <- train_data %>%
  mutate_if(is.numeric, scale)
```

```{r}
numeric_columns <- sapply(test_data, is.numeric)  # Identify numeric columns

# Scale the numeric columns while keeping column names
test_data[, numeric_columns] <- scale(test_data[, numeric_columns])

test_scaled_data <- test_data %>%
  mutate_if(is.numeric, scale)
```


```{r val_standardisation}
numeric_columns <- sapply(val_data, is.numeric)  # Identify numeric columns

# Scale the numeric columns while keeping column names
val_data[, numeric_columns] <- scale(val_data[, numeric_columns])

val_scaled_data <- val_data %>%
  mutate_if(is.numeric, scale)
```

## Logit Link Function

First we decided to perform stepAIC to get important covariates from the dataset.

```{r}
#Full model for backwards selection:
full_model <- glm(repay_fail ~ loan_amnt + term + int_rate + emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths + open_acc + pub_rec + revol_bal + revol_util + total_acc + credit_age_yrs, 
                 data = train_scaled_data, family = "binomial")

#Model with no variables present for forwards selection:
null_model <- glm(data = train_scaled_data,
    formula = repay_fail ~ 1,
    family = "binomial")

#Perform backward and forward selection:
backward_sel_model <- stepAIC(
  full_model, direction = "backward",trace = 0)
forward_sel_model <- stepAIC(
  null_model,
  scope = formula(full_model),
  direction = "forward",
  trace = 0) ## trace = 0 prevents automatic output of stepAIC function.
```

```{r}
formula(backward_sel_model)
```

create a model using the formula suggested by stepAIC:

```{r}
model1 <- glm(repay_fail ~ term + int_rate + emp_length + annual_inc + purpose + 
    inq_last_6mths + pub_rec + revol_bal + revol_util, data = train_scaled_data, family = binomial)
```

```{r}
AIC(model1)
```
Now creating a model a few interactions based on most relevance:

```{r}
model2 <- glm(repay_fail ~ term + int_rate + emp_length + annual_inc + purpose + 
    inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate + term * emp_length + term * annual_inc + term * purpose + term * pub_rec + int_rate * emp_length + int_rate * pub_rec + int_rate * annual_inc, data = train_scaled_data, family = binomial)
```

```{r}
AIC(model2)
```

The AIC of the new model is higher. We will now do an Anova test to confirm whether the addition of these interactions are adding value to the model compared to model without interactions:

```{r}
anova(model1, model2, test = "Chisq")
```

Based on the anova test, Model 2 is significantly better at explaining the variation in the dependent variable repay_fail compared to Model 1.

```{r}
model3 <- glm(repay_fail ~ term + int_rate + emp_length + annual_inc + purpose +
   inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate + term * emp_length + term * annual_inc + term * purpose + term * pub_rec,
     data = train_scaled_data, family = binomial)
```

```{r}
AIC(model3)
```

Model3 has the lowest AIC so far as compared to all other models we built.

```{r}
anova(model1, model3, test = "Chisq")
```

Based on the anova test, Model 3 is significantly better at explaining the variation in the dependent variable repay_fail compared to Model 1.

```{r}
summary(model3)
```

Now we select only the significant covariates from the above model to try and build a model with even lower AIC.

Fit logit model:

```{r fit_logit}
fit_logit <- glm(repay_fail ~ int_rate + emp_length + annual_inc + purpose + 
                   inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate,
                 data = train_scaled_data, family = binomial) #18156.43 

 AIC(fit_logit)
```
This model gives the lowest AIC, hence we chose to proceed with this model.

```{r summary_fit_logit}
summary(fit_logit)
```


```{r fit_logit_resid}
# simulate residuals from the model:
res = simulateResiduals(fit_logit)

# plot observed quantile versus expected quantile to assess distribution fit, and predicted value versus standardised residuals for unmodelled pattern in the residuals
plot(res)
```
DHARMA residual plots shows no sign of overdispersion: The distribution test (KS), Dispersion test and outlier test all are non-significant means there is evidence to suggest that the distribution of the simulated quantiles follows a uniform distribution. Additionally, the quantiles of the residuals (red lines) show a uniform pattern for predicted values (second plot).

```{r train_roc}
# ROC curve on scaled train data 
prob=predict(fit_logit,type=c("response"))
g <- roc(train_scaled_data$repay_fail ~ prob)
AUC <- g$auc
AUC
Gini <- 2*(AUC - 1/2)
Gini
plot(g)
```

```{r val_roc}
# ROC curve on scaled validation data
prob=predict(fit_logit,newdata=val_scaled_data,type=c("response"))
g <- roc(val_scaled_data$repay_fail ~ prob)
AUC <- g$auc
AUC
Gini <- 2*(AUC - 1/2)
Gini
plot(g)
```
The Gini score for the fitted logistic regression model is 0.410 on the training set and 0.395 on the validation set. Both of these scores are higher than those of the old model. Additionally, the ROC curve for the new model is positioned closer to the top-left corner compared to the old model. These findings collectively indicate that the new model is better than the old one.

## Probit Link Function

```{r fit_probit}
fit_probit <- glm(repay_fail ~ int_rate + emp_length + annual_inc + purpose + inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate,
    data = train_scaled_data, family = binomial("probit"))

AIC(fit_probit)
```

```{r probit_train_roc}
# Probit model: ROC curve on scaled train data 
prob=predict(fit_probit,type=c("response"))
g <- roc(train_scaled_data$repay_fail ~ prob)
AUC <- g$auc
AUC
Gini <- 2*(AUC - 1/2)
Gini
plot(g)
```

```{r probit_val_roc}
# Probit model: ROC curve on scaled validation data 
prob=predict(fit_probit,newdata = val_scaled_data,type=c("response"))
g <- roc(val_scaled_data$repay_fail ~ prob)
AUC <- g$auc
AUC
Gini <- 2*(AUC - 1/2)
Gini
plot(g)
```
While fit_probit model has lower AIC than fit_logit, but has slightly lower Gini score on the validation set.

## Cloglog Link Function

```{r fit_cloglog}
fit_cloglog <- glm(repay_fail ~ int_rate + emp_length + annual_inc + purpose + inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate,
    data = train_scaled_data, family = binomial("cloglog")) #18156.43

AIC(fit_cloglog)
```
fit_cloglog has highr AIC compare to both ligit_fit and probit_fit models.

Based on comparison conducted between three different link functions for our glm model, we decided to keep fit_logit as our final since it showed better performance.

## Calculate odds ratio and CI for the final model of GLM

```{r odds_ratio_GLM}
summary_logit <- summary(fit_logit)

parameter_estimates <- summary_logit$coef[, "Estimate"]
standard_errors <- summary_logit$coef[, "Std. Error"]
p_values <- summary_logit$coef[, "Pr(>|z|)"]

# Calculate the odds ratios by exponentiating the parameter estimate values
odds_ratios <- exp(parameter_estimates)

# Calculate the confidence intervals for the odds ratios
lower_ci <- exp(parameter_estimates - 1.96 * standard_errors)  # 1.96 corresponds to a 95% confidence interval
upper_ci <- exp(parameter_estimates + 1.96 * standard_errors)

model_results <- data.frame(
  Estimate = parameter_estimates,
  OddsRatio = odds_ratios,
  LowerCI = lower_ci,
  UpperCI = upper_ci,
  p_value = p_values
)

# Display results in a table
knitr::kable(
  model_results,
  col.names = c("Estimate", "Odds Ratio", "Lower Bound (2.5%)", "Upper Bound (97.5%)", "p-value")
)
```

Displaying only those variables whose p-value is less than 0.05:
```{r}
# Filter rows where p-value is less than 0.05
significant_results <- model_results[model_results$p_value < 0.05, ]

# Print the significant results
knitr::kable(
  significant_results,
  col.names = c("Estimate", "Odds Ratio", "Lower Bound (2.5%)", "Upper Bound (97.5%)", "p-value")
)
```

### Interpretations

Listed below are interpretations of the covariates of our final model:

- The intercept is statistically significant for individuals who have been employed for less than one year, are seeking a car loan, and have a 36-month loan term when all other predictors are equal to zero.

- For a one-unit increase in the interest rate, the odds of "repay_fail" occurring increase by a factor of approximately 1.57 (95% CI: 1.48-1.67). This is a significant effect with a very low p-value.

- For individuals with 1 to 10 plus years of employment, none of these employment lengths have a significant effect on the odds of "repay_fail" occurring compared to the reference category (less than one year"). Their odds ratios are close to 1, and their p-values are above 0.05, indicating that these employment lengths do not significantly impact the likelihood of loan repayment failure.

For individuals with missing or not specified employment length, or unemployed individual ("n/a"), For a one-unit increase in the emp_lengthn/a, the odds of "repay_fail" occurring increase by a factor of approximately 1.73 (95% CI: 1.37-2.19). This effect is statistically significant with a very low p-value, indicating that the absence of employment length information or for unemployed individual is associated with a higher likelihood of loan repayment failure.

- For a one-unit increase in annual income, the odds of "repay_fail" occurring decrease by a factor of approximately 0.71 (95% CI: 0.67-0.75). This is a significant effect with a very low p-value.

- For loans with the purpose of credit_card, debt_consolidation, educational, home_improvement, house, major_purchase, renewable_energy and wedding there is no significant effect on the odds of "repay_fail" occurring compared to the reference category (car). The p-value is above 0.05, indicating that these loan purposes do not significantly impact the likelihood of loan repayment failure.

For a one-unit increase in medical purpose, the odds of "repay_fail" occurring increase by a factor of approximately 1.83 (95% CI: 1.30-2.57).
For a one-unit increase in moving purpose, the odds of "repay_fail" occurring increase by a factor of approximately 1.65 (95% CI: 1.15-2.36).
For a one-unit increase in other purpose, the odds of "repay_fail" occurring increase by a factor of approximately 1.58 (95% CI: 1.24-2.00).
For a one-unit increase in small business purpose, the odds of "repay_fail" occurring increase by a factor of approximately 2.78 (95% CI: 2.15-3.59).  This effect is statistically significant with a very low p-value, indicating that the small business loan purpose is associated with a higher likelihood of loan repayment failure.
For a one-unit increase in vacation purpose, the odds of "repay_fail" occurring increase by a factor of approximately 1.56 (95% CI: 1.01-2.43). 

The specific loan purposes "medical," "moving," "other," and "small_business" are associated with a higher likelihood of loan repayment failure, and these effects are statistically significant. The effect of small business loan purpose is the most significant. "vacation" loans also have a statistically significant effect but to a lesser degree.
 
- For a one-unit increase in the number of inquiries in past 6 months, the odds of "repay_fail" occurring increase by a factor of approximately 1.24 (95% CI: 1.20-1.28). This is a significant effect with a very low p-value.

- For a one-unit increase in the number of derogatory public records, the odds of "repay_fail" occurring increase by a factor of approximately 1.06 (95% CI: 1.03-1.10). 

- For a one-unit increase in revolving balance, the odds of "repay_fail" occurring increase by a factor of approximately 1.10 (95% CI: 1.06-1.14).

- For a one-unit increase in revolving utilization, the odds of "repay_fail" occurring increase by a factor of approximately 1.12 (95% CI: 1.07-1.18).

- Having a 60-month loan term increases the odds of "repay_fail" occurring by a factor of approximately 1.67 (95% CI: 1.51-1.85) compared to the reference category (36 months loan term).

- The negative interaction term between interest rate and the 60-month term suggests that the effect of interest rate on "repay_fail" is less severe for loans with a 60-month term compared to the reference category 36 months terms. In other words, while interest rate still matters for loan repayment, its impact is less pronounced for loans with longer terms.
For a one-unit increase in interest rate, the odds of loan repayment failure decrease by a factor of approximately 0.9 (95% CI: 0.82-0.98) when the loan term is 60 months. The relationship between interest rate and loan repayment may differ depending on the term length, with the effect being less pronounced for longer-term loans.

## Odds ratio plot for the final model of GLM

```{r odds_ratio_plot_GLM}
plot_data <- data.frame(
  Variable = rownames(model_results),
  OddsRatio = model_results$OddsRatio,
  LowerCI = model_results$LowerCI,
  UpperCI = model_results$UpperCI
)

ggplot(plot_data, aes(x = OddsRatio, xmin = LowerCI, xmax = UpperCI, y = Variable)) +
  geom_point(size = 3, color="red") +
  geom_errorbarh(height = 0.2, color = "blue") +
  xlim(c(0, max(plot_data$UpperCI + 1))) +  
  theme_minimal() +
  labs(x = "Odds Ratio", y = "Variable") +
  ggtitle("Odds Ratios of Rapyment Failure for GLM")
```
Odds Ratio of Repayment Failure for GLM model including Confidence Interval ranges compared to baseline 36 months term, less than one year employment length, and loan purpose for car.

## Calculate accuracy and confusion matrix for the final model of GLM

```{r accuracy_confmatrix_GLM}
fit_logit <- ifelse(prob > 0.5, "1", "0")
fit_logit <- as.factor(fit_logit)

require(caret)
cm <- confusionMatrix(data=fit_logit,
                    reference=val_scaled_data$repay_fail,
                    positive="1")
cm
```

Included above is a confusion matrix, which depicts our model's true and false positive/negative rates. This can be used to better understand how well the model can predict new credit applications in terms of how accurately it classifies new applicants. The confusion matrix shows that out of 7,683 applicants, 6,504 of them will not default and have been accurately classified as so, and 21 of them will default and have been accurately classified as so. However, using our model, 1,138 applicants are falsely classified as defaulters when in fact, they would not have defaulted, and 20 applicants are falsely classified as not defaulting, when they are actually defaulters. 

## Part 1 Discussion

*** DISCUSS OVERALL FINDINGS

# Extended Credit Risk Model

Now, so far, we've been able to address management's concerns regarding the previous credit risk model and have presented a new model that performs much better. We are now going to address the second part of management's concerns, which includes using an extended version of the initial dataset. We will set out to answer whether accounting for variation in trends over jurisdiction or time changes or even improves performance benchmarks. Management has previously never considered whether credit risk changes between different states or over time, so we will investigate this.

We'll first take a look at the structure of the extended dataset.

```{r str_extended}
str(extended)
```

Next, we'll need to once again convert the relevant variables to factors. This time, this will include the new variables of zip_code, addr_state, and earliest_cr_line, in addition to the variables converted in the first part of this analysis:

```{r convert_vars_to_factors_ext}
extended$term <- as.factor(extended$term)
extended$emp_length <- as.factor(extended$emp_length)
extended$home_ownership <- as.factor(extended$home_ownership)
extended$verification_status <- as.factor(extended$verification_status)
extended$purpose <- as.factor(extended$purpose)
extended$repay_fail <- as.factor(extended$repay_fail)
extended$zip_code <- as.factor(extended$zip_code)
extended$addr_state <- as.factor(extended$addr_state)
extended$earliest_cr_line <- as.factor(extended$earliest_cr_line)
extended <- within(extended, location <- factor(as.factor(addr_state):zip_code)) # nest zip codes within states
```

```{r number of levels for each random effect}
nlevels(extended$zip_code)
nlevels(extended$addr_state)
nlevels(extended$location) # location includes zip nested within each state

nlevels(extended$earliest_cr_line)
```

Once again, we need to standardise the numeric variables in the dataset:

```{r standardisation_ext}
numeric_columns <- sapply(extended, is.numeric)  # Identify numeric columns

# Scale the numeric columns while keeping column names
extended[, numeric_columns] <- scale(extended[, numeric_columns])

extended_scaled <- extended %>%
  mutate_if(is.numeric, scale)
```

## Fit Models

Fit first model, with only the state as a random effect:

```{r extended_fit1}
fit_var_1 <- glmer(repay_fail ~ int_rate + emp_length + annual_inc + purpose +
                     inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate +
                     (1|addr_state), data = extended_scaled, family = binomial)
```

View first model summary with AIC equal to 30131.5:

```{r}
summary(fit_var_1)
```

Next, we'll add in another random effect of time, which is included as the variable "earliest_cr_line":

```{r}
fit_var_2 <- glmer(repay_fail ~ int_rate + emp_length + annual_inc + purpose +
                     inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate +
                     (1|addr_state) + (1|earliest_cr_line), data = extended_scaled, family = binomial)
```

Adding in time as a random effect caused a failed to converge warning. Thus, we will now increase the tolerance of this model to see if this helps:

```{r}
fit_var_3 <- glmer(repay_fail ~ int_rate + emp_length + annual_inc + purpose +
                     inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate +
                     (1|addr_state) + (1|earliest_cr_line), data = extended_scaled, family = binomial,
                   control = glmerControl(check.conv.grad = .makeCC("warning", tol = 3e-3, relTol = NULL)))
```

The model ran successfully, so we'll take a look at the summary:
AIC is 30126.5, and is lower than fit_var_1 which indicates the new model with both added random effects addr_state and earliest_cr_line is a better fit.

```{r}
summary(fit_var_3)
```

Finally, we'll add in one more random effect, which is zip code. Once again, we will fit this model with an increased tolerance, compared to the default tolerance:

```{r}
fit_var_4 <- glmer(repay_fail ~ int_rate + emp_length + annual_inc + purpose +
                     inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate +
                     (1|addr_state) + (1|earliest_cr_line) + (1|zip_code),
                   data = extended_scaled, family = binomial,
                   control = glmerControl(check.conv.grad = .makeCC("warning", tol = 9e-3, relTol = NULL)))
```

Now we can view this model summary:

```{r}
summary(fit_var_4)
```

Now, we'll try including just two random effects: earliest credit line (time), and then new location factor we created earlier, which combines states and zip codes in the form of state:zip.

```{r}
fit_var_5 <- glmer(repay_fail ~ int_rate + emp_length + annual_inc + purpose +
                     inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate +
                     (1|location) + (1|earliest_cr_line),
                   data = extended_scaled, family = binomial,
                   control = glmerControl(optimizer = "Nelder_Mead",
                                          check.conv.grad = .makeCC("warning", tol = 8e-3, relTol = NULL)))
```

```{r}
summary(fit_var_5)
```
```{r extended_roc}
par(mfrow = c(1, 2))
# ROC curve on scaled extended data  for fit_var_4
prob=predict(fit_var_4,type=c("response"))
g <- roc(extended_scaled$repay_fail ~ prob)
AUC <- g$auc
Gini <- 2*(AUC - 1/2)
Gini
plot(g, main = "ROC Curve for fit_var_4")

# ROC curve on scaled extended data  for fit_var_5
prob=predict(fit_var_5,type=c("response"))
g <- roc(extended_scaled$repay_fail ~ prob)
AUC <- g$auc
Gini <- 2*(AUC - 1/2)
Gini
plot(g, main = "ROC Curve for fit_var_5")
```

As the tolerance level for the above model was a bit high, we are now going to try to tune other paramaters to see if this allows for a lower tolerance.

```{r}
fit_var_6 <- glmer(repay_fail ~ int_rate + emp_length + annual_inc + purpose +
                     inq_last_6mths + pub_rec + revol_bal + revol_util + term * int_rate +
                     (1|addr_state) + (1|earliest_cr_line) + (1|zip_code),
                   data = extended_scaled, family = binomial,
                   control = glmerControl(optimizer = "Nelder_Mead", tolPwrss = 1e-3))
```


*** EXPLAIN

```{r runing_options - will be deleted}
# tuning options
# fit_var_2 <- update(fit_var_1, control = glmerControl(optimizer = "Nelder_Mead"))
# fit_var_3 <- update(fit_var_1, control = glmerControl(optimizer = "Nelder_Mead", tolPwrss = 1e-4))
# fit_var_4 <- update(fit_var_2, control = glmerControl(optCtrl=list(maxfun=2e4)))
```


## Analyse Performance Benchmarks with Extended Model

Analyse model with larger tolerance (all random effects separate):

```{r gof_fit_var_4}
# simulate residuals from the model:
res = simulateResiduals(fit_var_4)

# plot observed quantile versus expected quantile to assess distribution fit, and predicted value versus standardised residuals for unmodelled pattern in the residuals
plot(res)
```

Analyse model with random effects of (1) time and (2) zip code nested within state:

```{r gof_fit_var_5}
# simulate residuals from the model:
res = simulateResiduals(fit_var_5)

# plot observed quantile versus expected quantile to assess distribution fit, and predicted value versus standardised residuals for unmodelled pattern in the residuals
plot(res)
```

Does the model fit the data well?
```{r Goodness-of-fit}
plot(residuals(fit_var_5))
```
Look at the distribution of the random effects:
```{r res,include=TRUE}
res <- ranef(fit_var_5)
hist(res$earliest_cr_line[,1])

hist(res$location[,1])
```

## Calculate odds ratio and CI for the final GLMM model

Now, we will ...

```{r odds_ratio_GLMM}
summary_fit5 <- summary(fit_var_5)

parameter_estimates2 <- summary_fit5$coef[, "Estimate"]
standard_errors2 <- summary_fit5$coef[, "Std. Error"]
p_values2 <- summary_fit5$coef[, "Pr(>|z|)"]

# Calculate the odds ratios by exponentiating the parameter estimate values
odds_ratios2 <- exp(parameter_estimates2)

# Calculate the confidence intervals for the odds ratios
lower_ci2 <- exp(parameter_estimates2 - 1.96 * standard_errors2)  # 1.96 corresponds to a 95% confidence interval
upper_ci2 <- exp(parameter_estimates2 + 1.96 * standard_errors2)

model_results2 <- data.frame(
  Estimate = parameter_estimates2,
  OddsRatio = odds_ratios2,
  LowerCI = lower_ci2,
  UpperCI = upper_ci2,
  p_value = p_values2
)

# Display results in a table
knitr::kable(
  model_results2,
  col.names = c("Estimate", "Odds Ratio", "Lower Bound (2.5%)", "Upper Bound (97.5%)", "p-value")
)
```

Displaying only those variables whose p-value is less than 0.05:

```{r}
# Filter rows where p-value is less than 0.05
significant_results2 <- model_results2[model_results2$p_value < 0.05, ]

# Print the significant results
knitr::kable(
  significant_results2,
  col.names = c("Estimate", "Odds Ratio", "Lower Bound (2.5%)", "Upper Bound (97.5%)", "p-value")
)
```

## Odds ratio plot for the final model of GLMM

```{r odds_ratio_plot_GLM}
plot_data <- data.frame(
  Variable = rownames(model_results2),
  OddsRatio = model_results2$OddsRatio,
  LowerCI = model_results2$LowerCI,
  UpperCI = model_results2$UpperCI
)

ggplot(plot_data, aes(x = OddsRatio, xmin = LowerCI, xmax = UpperCI, y = Variable)) +
  geom_point(size = 3, color="red") +
  geom_errorbarh(height = 0.2, color = "blue") +
  xlim(c(0, max(plot_data$UpperCI + 1))) +  
  theme_minimal() +
  labs(x = "Odds Ratio", y = "Variable") +
  ggtitle("Odds Ratios of Rapyment Failure for GLMM")
```

### Interpretations

...

Newly significant covariates, after controlling for location and time:

* purpose == debt_consolidation

For a one-unit increase in the purpose of debt_consolidation, the odds of "repay_fail" occurring increase by a factor of approximately 1.24% (95% CI: 1.04-1.48). This effect is statistically significant with a very low p-value, indicating that loan taken out for the purpose of debt consolidation is associated with a higher likelihood of loan repayment failure.

* purpose == educational

For a one-unit increase in the purpose educational, the odds of "repay_fail" occurring increase by a factor of approximately 2.03% (95% CI: 1.50-2.76). This effect is statistically significant with a very low p-value.

* purpose == home_improvement

For a one-unit increase in the purpose home_improvement, the odds of "repay_fail" occurring increase by a factor of approximately 1.23% (95% CI: 1.00-1.50).

* purpose == renewable_energy

For a one-unit increase in the purpose renewable_energy, the odds of "repay_fail" occurring increase by a factor of approximately 1.82% (95% CI: 1.00-3.28).

## Part 2 Discussion

...

# Conclusion

...

## Caveats

* Extended model has a relatively high tolerance value, which could potentially be further reduced by tuning other model parameters
* We have only included a subset of the possible variables which may help to predict loan default
* We do not know the time frame of the data that is included, and it might be helpful to know this to ensure that we use recent data for model building and prediction tasks

## Future Directions

* You might want to explore other variables and interactions between variables that may help to predict loan default
* If provided with a validation dataset which included the extended variables of location and time, you could choose to use the extended model to predict loan default, and compare the extended model's prediction performance with that of the new model from part 1 of this analysis

# References

Listed below are the references we consulted to conduct this analysis.

[Control of Mixed Model Fitting](https://search.r-project.org/CRAN/refmans/lme4/html/lmerControl.html)

[Five Cs of Credit](https://www.investopedia.com/ask/answers/040115/what-most-important-c-five-cs-credit.asp)

[lme4 Convergence Warnings: Troubleshooting](https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html)

[lme4 Performance Tips](https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html#:~:text=choice%20of%20optimizer,may%20be%20worth%20a%20try.)

