---
title: "MXN600 Major Data Anaysis Project"
author: "Joanna Salerno, Pavan Asopa, Sevin Nejadi, Fernanda Martins Giuriati"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
# Introduction -- test---

The credit risk models our lending start up company uses are of the utmost importance to the functioning and ultimate success of the company. As we were recently acquired by a regional Australian bank, the success of our company also impacts the success of the bank. Recently, some of the bank's senior financial analysts have raised concerns about the credit risk models we have been using. They reviewed our models' performance benchmarks and feel as though our models are not suitable for use in a setting in which they are subject to strict regulatory requirements.

Thus, we have been tasked by the bank's management to rebuild our credit risk model from the ground up. As per management, our main objective is to use information known at the time of a loan application to build a model that predicts loan default. We will follow a standard statistical analysis process, which will be guided by the following questions:

1. How does this new model perform compared to the one used previously? How can it be expected to perform on new loan applications?
2. What are the important variables in this model and how do they compare to variables that are traditionally important for predicting credit risk in the banking sector?

Furthermore, management has consulted with an expert statistician, who has suggested we also account for variation in trends that may exist either between different jurisdictions or over time. The following questions will guide this second part of our analysis:

3. Can accounting for this variation (e.g., state/zip-code and time) improve performance benchmarks?
4. Are there any surprising differences in variables that are important for predicting credit risk?
5. Does credit risk change over time or between states? This is not something the bank has previously
investigated and results may inform modified loan policies in the future.

This report will document our entire analysis process, beginning with data exploration and cleaning, to model building and interpretations of our results.

# Setup

We will first load in the required libraries for our data exploration and analysis process.

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
library(GGally)
library(lme4)
library(lmerTest)
```

Next, we will load in our datasets. We have a total of 4:

(1) A training dataset that we will use to build and train our model
(2) A test dataset that we will use to test the fit of our model(s)
(3) A validation dataset that we will use to assess the performance of our model(s)
(4) An extended dataset that includes the necessary variables for us to account for variation such as location and time

```{r load_data}
train_data <- read.csv("benchmark_training_loan_data.csv")
test_data <- read.csv("benchmark_testing_loan_data.csv")
val_data <- read.csv("benchmark_validation_loan_data.csv")
extended <- read.csv("extendend_version_loan_data.csv")
```

# Exploratory Analysis

We will begin by exploring the available data to understand how each variable is distributed and to identify any potential data quality issues. We will also investigate the relationships between the different variables to see whether any variables are highly correlated with one another.

Note: For this exploration portion of our analysis, we will be using the training dataset.

We will first explore the training dataset to understand its structure and the variables it is comprised of.

```{r head_data}
head(train_data, 10)
```

```{r data_dim}
dim(train_data)
```

The training dataset contains a total of 23,052 observations of 19 variables.

```{r data_structure}
str(train_data)
```

Upon further investigation of the structure of the training data, we can see that 14 of the 19 variables are currently of numeric type, and the remaining 5 variables are characters. We will further investigate each variable to determine whether there are any outliers or missing data.

```{r data_summary}
summary(train_data)
```

Included above is the numeric distributions of each of the included variables.
*** FURTHER EXPLAIN ***

Now, we will briefly explore the data to see whether there are missing data for us to handle.

```{r sum_na}
sum(is.na(train_data))
```

Based on the above output, it appears as though, at least for numeric variables, this dataset does not contain any missing data in the form of NA values.

## Exploratory Plots

Below, we create a plot that depicts the correlation between the different numeric variables; it includes both correlation coefficients and visuals of the distribution of each included variable.

```{r ggpairs_plot}
# create ggpairs plot with the specified columns
ggpairs_plot <- ggpairs(train_data[, c("loan_amnt","int_rate", "annual_inc", "dti",
                                      "delinq_2yrs", "inq_last_6mths", "open_acc",
                                      "pub_rec", "revol_util", "revol_bal", "total_acc",
                                      "credit_age_yrs")],
                        progress = FALSE)

ggpairs_plot
```

Below, we create a correlation matrix of the numeric variables to help us understand the relationship between the different numeric variables.

```{r corr_matrix}
# correlation matrix
```

# New Credit Risk Model

```{r pois}
#Full model with all possible interactions for backwards selection:
full_interaction_model <- glm(data = train_data,
    formula = repay_fail ~ loan_amnt* term * int_rate * emp_length * home_ownership * annual_inc *
      verification_status * purpose * dti * delinq_2yrs * inq_last_6mths * open_acc * pub_rec *
      revol_bal * revol_util * total_acc * credit_age_yrs,
    family = "binomial")

#Model with no variables present for forwards selection:
null_model <- glm(data = train_data,
    formula = repay_fail ~ 1,
    family = "binomial")

#Perform backward and forward selection:
backward_sel_model <- stepAIC(
  full_interaction_model, direction = "backward",trace = 0)
forward_sel_model <- stepAIC(
  null_model,
  scope = formula(full_interaction_model), 
  direction = "forward",
  trace = 0) ## trace = 0 prevents automatic output of stepAIC function.
```


# Extended Credit Risk Model


# Conclusion


## Limitations


## Future Directions

